---
title: "Task Transcriptomics: RNA-seq Differential Expression Analysis"
author: "HackBio Internship"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# Global chunk options for all code chunks
knitr::opts_chunk$set(
  echo    = TRUE,     # show code in rendered output
  message = FALSE,    # suppress package loading messages
  warning = FALSE,    # suppress warnings
  fig.width  = 10,
  fig.height = 7
)
```

---

# 1. Background

This analysis uses a processed RNA-seq dataset comparing a **diseased cell line**
against a **diseased cell line treated with Compound X**. The key metrics are:

- **Log2FoldChange (Log2FC):** the magnitude of expression change between conditions.
  Positive = upregulated in treated; Negative = downregulated in treated.
- **p-value:** the statistical significance of the observed change.

**Thresholds used:**

| Direction      | Log2FC threshold | p-value threshold |
|----------------|-----------------|-------------------|
| Upregulated    | > 1             | < 0.01            |
| Downregulated  | < −1            | < 0.01            |
| Not significant| between −1 & 1  | ≥ 0.01            |

---

# 2. Load Packages

```{r packages}
# pacman auto-installs any missing package before loading
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,   # dplyr (data wrangling) + ggplot2 (visualization)
  ggplot2,     # explicit load for clarity
  ggrepel,     # smart non-overlapping gene labels on volcano plot
  scales       # axis formatting
)
```

---

# 3. Load & Inspect the Dataset

```{r load-data}
# ── Load the RNA-seq results from GitHub Gist ────────────────
# Dataset: DESeq2 output — diseased vs diseased + Compound X
# Columns: Gene, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj

url <- paste0(
  "https://gist.githubusercontent.com/stephenturner/",
  "806e31fce55a8b7175af/raw/",
  "1a507c4c3f9f1baaa3a69187223ff3d3050628d4/results.txt"
)

rnaseq <- read.table(url, header = TRUE, stringsAsFactors = FALSE)

# ── Quick overview ───────────────────────────────────────────
cat("=== Dataset Dimensions ===\n")
dim(rnaseq)                    # rows = genes, cols = metrics

cat("\n=== Column Names ===\n")
names(rnaseq)

cat("\n=== First Rows ===\n")
head(rnaseq)

cat("\n=== Summary Statistics ===\n")
summary(rnaseq)

cat("\n=== Missing Values per Column ===\n")
colSums(is.na(rnaseq))
```

> **What we observe:**
> The dataset contains expression statistics for thousands of genes.
> `log2FoldChange` ranges from large negative (strongly downregulated)
> to large positive (strongly upregulated) values.
> Some genes have `NA` in pvalue — these are genes with very low counts
> where DESeq2 could not compute reliable statistics. We will remove them
> before analysis.

---

# 4. Data Cleaning & Preparation

```{r data-prep}
# ── Remove rows with NA in pvalue or log2FoldChange ──────────
# NA values arise from genes with very low/zero counts in DESeq2
rnaseq_clean <- rnaseq %>%
  filter(!is.na(pvalue), !is.na(log2FoldChange))

cat(sprintf("Genes before cleaning : %d\n", nrow(rnaseq)))
cat(sprintf("Genes after cleaning  : %d\n", nrow(rnaseq_clean)))

# ── Add a -log10(pvalue) column ──────────────────────────────
# Volcano plots use -log10(pvalue) on the Y axis so that
# MORE significant genes appear HIGHER on the plot
rnaseq_clean <- rnaseq_clean %>%
  mutate(neg_log10_pval = -log10(pvalue))

# ── Classify each gene by expression status ──────────────────
# Using the task criteria:
#   Upregulated   : Log2FC >  1  AND pvalue < 0.01
#   Downregulated : Log2FC < -1  AND pvalue < 0.01
#   Not Significant: everything else

rnaseq_clean <- rnaseq_clean %>%
  mutate(Expression = case_when(
    log2FoldChange >  1 & pvalue < 0.01 ~ "Upregulated",
    log2FoldChange < -1 & pvalue < 0.01 ~ "Downregulated",
    TRUE                                 ~ "Not Significant"
  ))

# ── Count each category ──────────────────────────────────────
cat("\n=== Gene Expression Categories ===\n")
table(rnaseq_clean$Expression)
```

---

# 5. Volcano Plot

```{r volcano-plot, fig.height=8}
# ── What is a Volcano Plot? ──────────────────────────────────
# A volcano plot visualises differential expression by plotting:
#   X axis → Log2FoldChange (effect size: how much expression changed)
#   Y axis → -log10(pvalue) (significance: how confident we are)
#
# Genes appear as points:
#   TOP RIGHT   → highly significant AND strongly upregulated
#   TOP LEFT    → highly significant AND strongly downregulated
#   BOTTOM/CENTER → not significant or small fold change
#
# The "volcano" shape emerges from the V-like spread of points.

# ── Select top genes to label ────────────────────────────────
# Top 5 upregulated (highest Log2FC among significant)
top_up <- rnaseq_clean %>%
  filter(Expression == "Upregulated") %>%
  arrange(desc(log2FoldChange)) %>%
  head(5)

# Top 5 downregulated (lowest/most negative Log2FC among significant)
top_down <- rnaseq_clean %>%
  filter(Expression == "Downregulated") %>%
  arrange(log2FoldChange) %>%
  head(5)

# Combine for labeling
top_genes <- bind_rows(top_up, top_down)

# ── Build the volcano plot ───────────────────────────────────
volcano_plot <- ggplot(rnaseq_clean,
                       aes(x     = log2FoldChange,
                           y     = neg_log10_pval,
                           color = Expression)) +

  # All gene points
  geom_point(alpha = 0.5, size = 1.5) +

  # Threshold lines
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "grey30", linewidth = 0.7) +
  geom_hline(yintercept = -log10(0.01),
             linetype = "dashed", color = "grey30", linewidth = 0.7) +

  # Gene labels for top hits — ggrepel avoids overlapping text
  ggrepel::geom_label_repel(
    data          = top_genes,
    aes(label     = Gene),
    size          = 3.2,
    fontface      = "bold",
    box.padding   = 0.4,
    point.padding = 0.3,
    max.overlaps  = 20,
    show.legend   = FALSE
  ) +

  # Color scheme: red = up, blue = down, grey = not significant
  scale_color_manual(
    values = c(
      "Upregulated"     = "#B71C1C",
      "Downregulated"   = "#1565C0",
      "Not Significant" = "grey60"
    )
  ) +

  # Axis labels and title
  labs(
    title    = "Volcano Plot — Differential Gene Expression",
    subtitle = "Diseased Cell Line vs Diseased + Compound X Treatment",
    x        = "Log₂ Fold Change  (←  Downregulated  |  Upregulated  →)",
    y        = "-log₁₀(p-value)  [higher = more significant]",
    color    = "Expression Status",
    caption  = paste0(
      "Thresholds: |Log2FC| > 1 and p-value < 0.01\n",
      "Source: Turner (2015) RNA-seq results via GitHub Gist"
    )
  ) +

  # Annotate quadrant counts
  annotate("text", x =  max(rnaseq_clean$log2FoldChange) * 0.8,
           y = max(rnaseq_clean$neg_log10_pval) * 0.95,
           label = paste0("Up: ", nrow(top_up %>%
                            {rnaseq_clean[rnaseq_clean$Expression=="Upregulated",]})),
           color = "#B71C1C", fontface = "bold", size = 4) +

  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "grey50", size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(volcano_plot)

# ── OBSERVATION ──────────────────────────────────────────────
# The volcano plot shows a clear separation between:
# - Upregulated genes (top-right): these are ACTIVATED by Compound X treatment
# - Downregulated genes (top-left): these are SUPPRESSED by Compound X treatment
# - The bulk of genes cluster near the centre/bottom: no meaningful change
# The shape resembles a volcano because most genes are not significant
# (low -log10 p-value) and only a subset show both large fold change
# AND high significance simultaneously.
```

---

# 6. Upregulated Genes (Log2FC > 1 and p-value < 0.01)

```{r upregulated}
# ── Filter upregulated genes ─────────────────────────────────
# These genes are MORE expressed in the treated condition
# → Compound X activates or induces their expression

upregulated <- rnaseq_clean %>%
  filter(Expression == "Upregulated") %>%
  arrange(desc(log2FoldChange)) %>%
  select(Gene, log2FoldChange, pvalue, padj, baseMean)

cat(sprintf("Total upregulated genes: %d\n\n", nrow(upregulated)))

# ── Top 5 most upregulated genes ─────────────────────────────
cat("=== TOP 5 UPREGULATED GENES ===\n")
print(head(upregulated, 5))
```

> **Interpretation:**
> Upregulated genes are those **switched ON** (or increased) in response to
> Compound X treatment in the diseased cell line. A Log2FC > 1 means
> expression at least **doubled** (2¹ = 2-fold increase) in treated cells.
> These genes may represent the compound's mechanism of therapeutic action —
> pathways it activates to counteract the disease state.

---

# 7. Downregulated Genes (Log2FC < -1 and p-value < 0.01)

```{r downregulated}
# ── Filter downregulated genes ───────────────────────────────
# These genes are LESS expressed in the treated condition
# → Compound X suppresses or silences their expression

downregulated <- rnaseq_clean %>%
  filter(Expression == "Downregulated") %>%
  arrange(log2FoldChange) %>%       # most negative first
  select(Gene, log2FoldChange, pvalue, padj, baseMean)

cat(sprintf("Total downregulated genes: %d\n\n", nrow(downregulated)))

# ── Top 5 most downregulated genes ───────────────────────────
cat("=== TOP 5 DOWNREGULATED GENES ===\n")
print(head(downregulated, 5))
```

> **Interpretation:**
> Downregulated genes are those **switched OFF** (or reduced) by Compound X.
> A Log2FC < −1 means expression at least **halved** (2⁻¹ = 2-fold decrease).
> These genes may represent disease-promoting pathways that Compound X
> successfully suppresses — making them strong candidates for therapeutic targets.

---

# 8. Function of Top 5 Upregulated Genes (from GeneCards)

```{r top-upregulated-functions}
# ── Store the top 5 upregulated gene names ───────────────────
top5_up <- head(upregulated$Gene, 5)
cat("Top 5 upregulated genes:\n")
print(top5_up)
```

Based on [GeneCards](https://www.genecards.org), here are the functions of the
top 5 upregulated genes:

---

### 1. `DTHD1` — Death Domain Containing 1

- **Full name:** Death Domain Containing 1
- **Function:** Encodes a protein containing a **death domain** — a protein
  interaction motif involved in **apoptosis (programmed cell death)** signalling.
  Death domain proteins participate in the assembly of signalling complexes that
  activate caspase cascades leading to cell death.
- **Relevance:** Upregulation of DTHD1 by Compound X suggests the treatment
  may be **re-activating apoptotic pathways** in diseased cells — a key mechanism
  in cancer therapy where diseased cells evade normal cell death.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=DTHD1

---

### 2. `EMILIN2` — Elastin Microfibril Interfacer 2

- **Full name:** Elastin Microfibril Interface Located Protein 2
- **Function:** A secreted glycoprotein of the **extracellular matrix (ECM)**
  that plays roles in:
  - **Elastic fibre assembly** — structural component of connective tissue
  - **Apoptosis regulation** — EMILIN2 activates the extrinsic apoptotic
    pathway via caspase-8 and caspase-3
  - **Angiogenesis inhibition** — suppresses tumour vascularisation
  - **Tumour suppression** — reduced expression is observed in several cancers
- **Relevance:** Upregulation suggests Compound X may be **restoring tumour
  suppressive ECM signalling** and **pro-apoptotic signals** in diseased cells.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=EMILIN2

---

### 3. `ACTC1` — Actin Alpha Cardiac Muscle 1

- **Full name:** Actin, Alpha Cardiac Muscle 1
- **Function:** Encodes the **alpha-cardiac actin** isoform — a core structural
  protein of the **sarcomere** (the contractile unit of cardiac and skeletal muscle).
  Key roles include:
  - **Muscle contraction** — forms thin filaments that interact with myosin
  - **Cardiac development** — expressed during heart morphogenesis
  - **Disease relevance:** Mutations cause **hypertrophic cardiomyopathy** and
    **dilated cardiomyopathy**
- **Relevance:** Upregulation in this context may indicate Compound X
  **stimulates muscle repair or cytoskeletal reorganisation** in the diseased cells.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=ACTC1

---

### 4. `LOC100130155` — Uncharacterised Locus

- **Full name:** Uncharacterised locus / predicted gene
- **Function:** This is a **predicted/uncharacterized open reading frame** with
  no confirmed protein function at the time of the original study. It may encode:
  - A **long non-coding RNA (lncRNA)**
  - A **novel protein** with functions yet to be experimentally validated
- **Relevance:** Its significant upregulation suggests it may play an
  **undiscovered regulatory role** in the cellular response to Compound X.
  Such loci are active areas of functional genomics research.

---

### 5. `SPDYE7P` — Speedy/RINGO Cell Cycle Regulator Family Member E7, Pseudogene

- **Full name:** Speedy/RINGO Cell Cycle Regulator Family Member E7, Pseudogene
- **Function:** This is a **pseudogene** derived from the Speedy/RINGO family,
  which normally functions as:
  - **Cell cycle regulators** — activate CDK (cyclin-dependent kinase) complexes
    independently of cyclins
  - **Accelerators of G1/S and G2/M transitions** in the cell cycle
  - Although classified as a pseudogene, recent evidence shows some pseudogenes
    produce **functional RNAs** that regulate expression of parent genes
- **Relevance:** Its upregulation may reflect **Compound X modulating cell cycle
  progression** in the diseased cell line.

---

# 9. Function of Top 5 Downregulated Genes (from GeneCards)

```{r top-downregulated-functions}
# ── Store the top 5 downregulated gene names ─────────────────
top5_down <- head(downregulated$Gene, 5)
cat("Top 5 downregulated genes:\n")
print(top5_down)
```

Based on [GeneCards](https://www.genecards.org), here are the functions of the
top 5 downregulated genes:

---

### 1. `CRISPLD2` — Cysteine Rich Secretory Protein LCCL Domain Containing 2

- **Full name:** Cysteine Rich Secretory Protein LCCL Domain Containing 2
- **Function:** Encodes a secreted extracellular matrix protein involved in:
  - **Lipopolysaccharide (LPS) binding** — neutralises bacterial endotoxins
  - **Anti-inflammatory regulation** — suppresses inflammatory cytokine production
  - **Palate and craniofacial development** — linked to cleft palate
  - **Lung development** — expressed in airway smooth muscle
- **Relevance:** Downregulation by Compound X may indicate **reduction of
  inflammatory signalling components** in the diseased cell line.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=CRISPLD2

---

### 2. `MXRA5` — Matrix Remodelling Associated 5

- **Full name:** Matrix Remodelling Associated Protein 5
- **Function:** A secreted glycoprotein involved in:
  - **Extracellular matrix (ECM) remodelling** — regulates cell adhesion
  - **Tumour progression** — overexpressed in glioma and associated with
    **poor prognosis**, immune checkpoint upregulation, and macrophage infiltration
  - **Mesenchymal transition** — marker of the mesenchymal subtype of GBM
- **Relevance:** Downregulation is a **positive therapeutic signal** — Compound X
  may be **suppressing tumour-promoting ECM remodelling** and reducing
  the immunosuppressive tumour microenvironment.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=MXRA5

---

### 3. `ITGBL1` — Integrin Subunit Beta Like 1

- **Full name:** Integrin Beta-Like Protein 1
- **Function:** A protein structurally related to integrin beta subunits that:
  - **Modulates integrin-mediated cell adhesion** to ECM components
  - **Promotes cancer metastasis** — upregulated in colorectal and ovarian cancers,
    facilitates epithelial-mesenchymal transition (EMT)
  - **Activates TGF-β signalling** — a key pro-fibrotic and pro-tumour pathway
  - **Inhibits Wnt signalling** in some contexts
- **Relevance:** Its downregulation suggests Compound X may be
  **reducing cancer cell invasion and metastatic potential** by disrupting
  integrin-mediated adhesion signalling.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=ITGBL1

---

### 4. `HBB` — Haemoglobin Subunit Beta

- **Full name:** Haemoglobin Subunit Beta (Beta-Globin)
- **Function:** Encodes the **beta chain of adult haemoglobin**, the
  oxygen-carrying protein in red blood cells. Key roles:
  - **Oxygen transport** — binds O₂ cooperatively in a tetrameric complex
    with alpha-globin subunits
  - **Disease relevance:** Mutations cause **sickle cell anaemia** (HbS),
    **beta-thalassaemia**, and other haemoglobinopathies
  - Also expressed at low levels in non-erythroid tissues where it may
    function in **oxidative stress protection**
- **Relevance:** HBB downregulation in a diseased cell line context may
  indicate **altered oxygen metabolism or haem biosynthesis** in cells
  responding to Compound X.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=HBB

---

### 5. `APCDD1L` — APC Down-Regulated 1 Like

- **Full name:** Adenomatous Polyposis Coli Down-Regulated 1 Like
- **Function:** A transmembrane protein that:
  - **Inhibits the Wnt signalling pathway** — by promoting degradation of
    Wnt pathway components
  - **Regulates hair follicle development** — mutations cause
    Hereditary Hypotrichosis Simplex (hair loss disorder)
  - **Tumour suppressor activity** — downregulated in colorectal cancers
    where Wnt signalling is hyperactivated
- **Relevance:** Downregulation of APCDD1L appears counterintuitive for a
  tumour suppressor, but may reflect **cell-type-specific Wnt pathway
  rebalancing** as part of Compound X's mechanism of action.
- **GeneCards:** https://www.genecards.org/cgi-bin/carddisp.pl?gene=APCDD1L

---

# 10. Summary Statistics & Final Table

```{r summary-tables}
# ── Full summary of differentially expressed genes ───────────
cat("=== DIFFERENTIAL EXPRESSION SUMMARY ===\n")
cat(sprintf("Total genes analysed       : %d\n", nrow(rnaseq_clean)))
cat(sprintf("Upregulated   (FC>1, p<0.01): %d\n",
            sum(rnaseq_clean$Expression == "Upregulated")))
cat(sprintf("Downregulated (FC<-1, p<0.01): %d\n",
            sum(rnaseq_clean$Expression == "Downregulated")))
cat(sprintf("Not significant             : %d\n",
            sum(rnaseq_clean$Expression == "Not Significant")))

# ── Top 5 up and down in a combined table ────────────────────
cat("\n=== TOP 5 UPREGULATED ===\n")
print(head(upregulated %>%
             select(Gene, log2FoldChange, pvalue), 5))

cat("\n=== TOP 5 DOWNREGULATED ===\n")
print(head(downregulated %>%
             select(Gene, log2FoldChange, pvalue), 5))
```

---

# 11. Enhanced Volcano Plot with All Labels

```{r enhanced-volcano, fig.height=9}
# ── Final enhanced volcano with both up and down top genes ───
ggplot(rnaseq_clean,
       aes(x     = log2FoldChange,
           y     = neg_log10_pval,
           color = Expression)) +
  geom_point(alpha = 0.45, size = 1.4) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "grey20", linewidth = 0.6) +
  geom_hline(yintercept = -log10(0.01),
             linetype = "dashed", color = "grey20", linewidth = 0.6) +
  ggrepel::geom_label_repel(
    data        = top_genes,
    aes(label   = Gene),
    size        = 3.2,
    fontface    = "bold",
    fill        = "white",
    alpha       = 0.85,
    box.padding = 0.5,
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Upregulated"     = "#C62828",
      "Downregulated"   = "#1565C0",
      "Not Significant" = "grey70"
    )
  ) +
  # Annotate counts in each quadrant
  annotate("text",
           x =  max(rnaseq_clean$log2FoldChange, na.rm=TRUE) * 0.7,
           y =  max(rnaseq_clean$neg_log10_pval, na.rm=TRUE) * 0.92,
           label = paste0(sum(rnaseq_clean$Expression=="Upregulated"),
                          " Upregulated"),
           color = "#C62828", fontface = "bold", size = 4) +
  annotate("text",
           x =  min(rnaseq_clean$log2FoldChange, na.rm=TRUE) * 0.7,
           y =  max(rnaseq_clean$neg_log10_pval, na.rm=TRUE) * 0.92,
           label = paste0(sum(rnaseq_clean$Expression=="Downregulated"),
                          " Downregulated"),
           color = "#1565C0", fontface = "bold", size = 4) +
  labs(
    title    = "Volcano Plot — RNA-seq Differential Expression",
    subtitle = "Diseased Cell Line vs Diseased + Compound X  |  Top 5 Up & Down genes labelled",
    x        = "Log₂ Fold Change",
    y        = "-log₁₀(p-value)",
    color    = "Expression Status",
    caption  = "Thresholds: |Log2FC| > 1 and p-value < 0.01  |  Source: Turner (2015)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

---

# 12. Overall Biological Conclusions

> **Upregulated genes** — Compound X activates genes involved in:
> - **Apoptosis** (`DTHD1`, `EMILIN2`) → the treatment is re-enabling
>   programmed cell death in diseased cells that had evaded it
> - **Structural remodelling** (`ACTC1`) → cytoskeletal reorganisation
>   as part of the cellular response
> - **Cell cycle regulation** (`SPDYE7P`) → potential modulation of
>   proliferative capacity

> **Downregulated genes** — Compound X suppresses genes involved in:
> - **Tumour ECM remodelling** (`MXRA5`, `ITGBL1`) → reduced invasion
>   and metastatic potential
> - **Inflammation** (`CRISPLD2`) → dampening of inflammatory signalling
> - **Wnt pathway** (`APCDD1L`) → pathway rebalancing in treated cells
> - **Oxygen metabolism** (`HBB`) → altered metabolic state

> **Overall conclusion:**
> Compound X appears to act through a **dual mechanism** — simultaneously
> activating pro-apoptotic and structural repair pathways while suppressing
> tumour-promoting ECM remodelling and inflammatory signals. This pattern
> is consistent with a compound that **re-sensitises diseased cells to
> normal regulatory control** — a hallmark of effective therapeutic intervention.

---

*End of Task 2.6 — Transcriptomics RNA-seq Differential Expression Analysis*
