---
title: "Task Transcriptomics: RNA-seq Differential Expression Analysis"
author: "HackBio Internship"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# Global chunk options for all code chunks
knitr::opts_chunk$set(
  echo    = TRUE,     # show code in rendered output
  message = FALSE,    # suppress package loading messages
  warning = FALSE,    # suppress warnings
  fig.width  = 10,
  fig.height = 7
)
```

---

# 1. Background

This analysis uses a processed RNA-seq dataset comparing a **diseased cell line**
against a **diseased cell line treated with Compound X**. The key metrics are:

- **Log2FoldChange (Log2FC):** the magnitude of expression change between conditions.
  Positive = upregulated in treated; Negative = downregulated in treated.
- **p-value:** the statistical significance of the observed change.

**Thresholds used:**

| Direction      | Log2FC threshold | p-value threshold |
|----------------|-----------------|-------------------|
| Upregulated    | > 1             | < 0.01            |
| Downregulated  | < −1            | < 0.01            |
| Not significant| between −1 & 1  | ≥ 0.01            |

---

# 2. Load Packages

```{r packages}
# pacman auto-installs any missing package before loading
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,   # dplyr (data wrangling) + ggplot2 (visualization)
  ggplot2,     # explicit load for clarity
  ggrepel,     # smart non-overlapping gene labels on volcano plot
  scales       # axis formatting
)
```

---

# 3. Load & Inspect the Dataset

```{r load-data}
# ── Load the RNA-seq results from GitHub Gist ────────────────
# Dataset: DESeq2 output — diseased vs diseased + Compound X
# Columns: Gene, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj

url <- paste0(
  "https://gist.githubusercontent.com/stephenturner/",
  "806e31fce55a8b7175af/raw/",
  "1a507c4c3f9f1baaa3a69187223ff3d3050628d4/results.txt"
)

Raw_rnaseq <- read.table(url, header = TRUE, stringsAsFactors = FALSE)
rnaseq <- Raw_rnaseq # always conserve the Raw data 

# ── Quick overview ───────────────────────────────────────────
cat("=== Dataset Dimensions ===\n")
dim(rnaseq)                    # rows = genes, cols = metrics

cat("\n=== Column Names ===\n")
names(rnaseq)

cat("\n=== First Rows ===\n")
head(rnaseq)

cat("\n=== Summary Statistics ===\n")
summary(rnaseq)

cat("\n=== Missing Values per Column ===\n")
colSums(is.na(rnaseq)) # no misisng values in this dataset

# save raw data in the computer
write.csv(rnaseq, "Raw_rnaseq.csv", row.names = FALSE)

```

> **What we observe:**
> The dataset contains expression statistics for thousands of genes.
> `log2FoldChange` ranges from large negative (strongly downregulated)
> to large positive (strongly upregulated) values.
> If we have genes with `NA` in pvalue — these are genes with very low counts
> where DESeq2 could not compute reliable statistics. We will remove them
> before analysis.

---

# 4. Data Cleaning & Preparation

```{r data-prep}
# ── Remove rows with NA in pvalue or log2FoldChange ──────────
# NA values arise from genes with very low/zero counts in DESeq2
rnaseq_clean <- rnaseq %>%
  filter(!is.na(pvalue), !is.na(log2FoldChange))

cat(sprintf("Genes before cleaning : %d\n", nrow(rnaseq))) # there are no genes with NA values
cat(sprintf("Genes after cleaning  : %d\n", nrow(rnaseq_clean)))

# ── Add a -log10(pvalue) column ──────────────────────────────
# Volcano plots use -log10(pvalue) on the Y axis so that
# MORE significant genes appear HIGHER on the plot
rnaseq_clean <- rnaseq_clean %>%
  mutate(neg_log10_pval = -log10(pvalue))

# ── Classify each gene by expression status ──────────────────
# Using the task criteria:
#   Upregulated   : Log2FC >  1  AND pvalue < 0.01
#   Downregulated : Log2FC < -1  AND pvalue < 0.01
#   Not Significant: everything else

rnaseq_clean <- rnaseq_clean %>%
  mutate(Expression = case_when(
    log2FoldChange >  1 & pvalue < 0.01 ~ "Upregulated",
    log2FoldChange < -1 & pvalue < 0.01 ~ "Downregulated",
    TRUE                                 ~ "Not Significant" # considering the criteria from the exercice
  ))

# ── Count each category ──────────────────────────────────────
cat("\n=== Gene Expression Categories ===\n")
table(rnaseq_clean$Expression)
```

---

# 5. Volcano Plot

```{r volcano-plot, fig.height=8}
# ── What is a Volcano Plot? ──────────────────────────────────
# A volcano plot visualises differential expression by plotting:
#   X axis → Log2FoldChange (effect size: how much expression changed)
#   Y axis → -log10(pvalue) (significance: how confident we are)
#
# Genes appear as points:
#   TOP RIGHT   → highly significant AND strongly upregulated
#   TOP LEFT    → highly significant AND strongly downregulated
#   BOTTOM/CENTER → not significant or small fold change
#
# The "volcano" shape emerges from the V-like spread of points.

# ── Select top genes to label ────────────────────────────────
# Top 5 upregulated (highest Log2FC among significant)
top_up <- rnaseq_clean %>%
  filter(Expression == "Upregulated") %>%
  arrange(desc(log2FoldChange)) %>%
  head(5)
top_up

# Top 5 downregulated (lowest/most negative Log2FC among significant)
top_down <- rnaseq_clean %>%
  filter(Expression == "Downregulated") %>%
  arrange(log2FoldChange) %>%
  head(5)
top_down


# Combine for labeling
top_genes <- bind_rows(top_up, top_down)
top_genes

# ── Build the volcano plot ───────────────────────────────────
volcano_plot <- ggplot(rnaseq_clean,
                       aes(x     = log2FoldChange,
                           y     = neg_log10_pval,
                           color = Expression)) +

  # All gene points
  geom_point(alpha = 0.5, size = 1.5) +

  # Threshold lines
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "grey30", linewidth = 0.7) +
  geom_hline(yintercept = -log10(0.01),
             linetype = "dashed", color = "grey30", linewidth = 0.7) +

  # Gene labels for top hits — ggrepel avoids overlapping text
  ggrepel::geom_label_repel(
    data          = top_genes,
    aes(label     = Gene),
    size          = 3.2,
    fontface      = "bold",
    box.padding   = 0.4,
    point.padding = 0.3,
    max.overlaps  = 20,
    show.legend   = FALSE
  ) +

  # Color scheme: red = up, blue = down, grey = not significant
  scale_color_manual(
    values = c(
      "Upregulated"     = "#B71C1C",
      "Downregulated"   = "#1565C0",
      "Not Significant" = "grey60"
    )
  ) +

  # Axis labels and title
  labs(
    title    = "Volcano Plot — Differential Gene Expression",
    subtitle = "Diseased Cell Line vs Diseased + Compound X Treatment",
    x        = "Log₂ Fold Change  (←  Downregulated  |  Upregulated  →)",
    y        = "-log₁₀(p-value)  [higher = more significant]",
    color    = "Expression Status",
    caption  = paste0(
      "Thresholds: |Log2FC| > 1 and p-value < 0.01\n",
      "Source: Turner (2015) RNA-seq results via GitHub Gist"
    )
  ) +

  # Annotate quadrant counts
  annotate("text", x =  max(rnaseq_clean$log2FoldChange) * 0.8,
           y = max(rnaseq_clean$neg_log10_pval) * 0.95,
           label = paste0("Up: ", nrow(top_up %>%
                            {rnaseq_clean[rnaseq_clean$Expression=="Upregulated",]})),
           color = "#B71C1C", fontface = "bold", size = 4) +

  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "grey50", size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(volcano_plot)

ggsave("Volcano.png", width = 12, height = 8, dpi = 150)
cat("✔ Saved: Volcano Plot.png\n")

# ── OBSERVATION ──────────────────────────────────────────────
# The volcano plot shows a clear separation between:
# - Upregulated genes (top-right): these are ACTIVATED by Compound X treatment
# - Downregulated genes (top-left): these are SUPPRESSED by Compound X treatment
# - The bulk of genes cluster near the centre/bottom: no meaningful change
# The shape resembles a volcano because most genes are not significant
# (low -log10 p-value) and only a subset show both large fold change
# AND high significance simultaneously.
```

---

# 6. Upregulated Genes (Log2FC > 1 and p-value < 0.01)

```{r upregulated}
# ── Filter upregulated genes ─────────────────────────────────
# These genes are MORE expressed in the treated condition
# → Compound X activates or induces their expression

upregulated <- rnaseq_clean %>%
  filter(Expression == "Upregulated") %>%
  arrange(desc(log2FoldChange)) %>%
  select(Gene, log2FoldChange, pvalue, padj, neg_log10_pval)

cat(sprintf("Total upregulated genes: %d\n\n", nrow(upregulated)))

# ── Top 5 most upregulated genes ─────────────────────────────
cat("=== TOP 5 UPREGULATED GENES ===\n")
print(head(upregulated, 5))
```

> **Interpretation:**
> Upregulated genes are those **switched ON** (or increased) in response to
> Compound X treatment in the diseased cell line. A Log2FC > 1 means
> expression at least **doubled** (2¹ = 2-fold increase) in treated cells.
> These genes may represent the compound's mechanism of therapeutic action —
> pathways it activates to counteract the disease state.

---

# 7. Downregulated Genes (Log2FC < -1 and p-value < 0.01)

```{r downregulated}
# ── Filter downregulated genes ───────────────────────────────
# These genes are LESS expressed in the treated condition
# → Compound X suppresses or silences their expression

downregulated <- rnaseq_clean %>%
  filter(Expression == "Downregulated") %>%
  arrange(log2FoldChange) %>%       # most negative first
  select(Gene, log2FoldChange, pvalue, padj, neg_log10_pval)

cat(sprintf("Total downregulated genes: %d\n\n", nrow(downregulated)))

# ── Top 5 most downregulated genes ───────────────────────────
cat("=== TOP 5 DOWNREGULATED GENES ===\n")
print(head(downregulated, 5))
```

> **Interpretation:**
> Downregulated genes are those **switched OFF** (or reduced) by Compound X.
> A Log2FC < −1 means expression at least **halved** (2⁻¹ = 2-fold decrease).
> These genes may represent disease-promoting pathways that Compound X
> successfully suppresses — making them strong candidates for therapeutic targets.

---

# 8. Function of Top 5 Upregulated Genes (from GeneCards)

```{r top-upregulated-functions}
# ── Store the top 5 upregulated gene names ───────────────────
top5_up <- head(upregulated$Gene, 5)
cat("Top 5 upregulated genes:\n")
print(top5_up)
```

Based on [GeneCards](https://www.genecards.org), here are the functions of the
top 5 upregulated genes:

---

### 1. `DTHD1` — Death Domain Containing 1

- **Full name:** Death Domain Containing 1
- **Function:** Encodes a protein containing a **death domain** — a protein
  interaction motif involved in **apoptosis (programmed cell death)** signalling.
  Death domain proteins participate in the assembly of signalling complexes that
  activate caspase cascades leading to cell death.
- **Relevance:** Upregulation of DTHD1 by Compound X suggests the treatment
  may be **re-activating apoptotic pathways** in diseased cells — a key mechanism
  in cancer therapy where diseased cells evade normal cell death.
- **GeneCards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=DTHD1>

---

### 2. `EMILIN2` — Elastin Microfibril Interfacer 2

- **Full name:** Elastin Microfibril Interface Located Protein 2
- **Function:** A secreted glycoprotein of the **extracellular matrix (ECM)**
  that plays roles in:
  - **Elastic fibre assembly** — structural component of connective tissue
  - **Apoptosis regulation** — EMILIN2 activates the extrinsic apoptotic
    pathway via caspase-8 and caspase-3
  - **Angiogenesis inhibition** — suppresses tumour vascularisation
  - **Tumour suppression** — reduced expression is observed in several cancers
- **Relevance:** Upregulation suggests Compound X may be **restoring tumour
  suppressive ECM signalling** and **pro-apoptotic signals** in diseased cells.
- **GeneCards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=EMILIN2>

---

### 3. `PI16` - Peptidaase Inhibitor 16

- **Full name:** Peptidase Inhibitor 16
- **Function:** Predicted to enable peptidase inhibitor activity. Predicted to act upstream of or within negative regulation of cell growth involved in cardiac muscle cell development. Predicted to be located in extracellular region. Predicted to be active in extracellular space
- **Genecards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=PI16>
---

### 4. `C4orf45` - Gene - Sperm Microtubule Inner Protein 2

-**Full name:** Gene - Sperm Microtubule Inner Protein 2
- **Functions:** PMIP2 (Sperm Microtubule Inner Protein 2) is a Protein Coding gene. Diseases associated with SPMIP2 include Hyperekplexia
- **Genecards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=SPMIP2>

---

### 5. `FAM180B` — Family With Sequence Similarity 180 Member B

- **Full name:** Family With Sequence Similarity 180 Member B
- **Function:** Predicted to be located in extracellular region. FAM180B (Family With Sequence Similarity 180 Member B) is a Protein Coding gene. Diseases associated with FAM180B include Borderline Leprosy and Mosaic Variegated Aneuploidy Syndrome.
-**Genecards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=FAM180B>

---

# 9. Function of Top 5 Downregulated Genes (from GeneCards)

```{r top-downregulated-functions}
# ── Store the top 5 downregulated gene names ─────────────────
top5_down <- head(downregulated$Gene, 5)
cat("Top 5 downregulated genes:\n")
print(top5_down)
```

Based on [GeneCards](https://www.genecards.org), here are the functions of the
top 5 downregulated genes:

---

### 1. `TBX5` - T-Box Transcription Factor 

- **Full name:** T-Box Transcription Factor 
- **Function:** This gene is a member of a phylogenetically conserved family of genes that share a common DNA-binding domain, the T-box. T-box genes encode transcription factors involved in the regulation of developmental processes. This gene is closely linked to related family member T-box 3 (ulnar mammary syndrome) on human chromosome 12. The encoded protein may play a role in heart development and specification of limb identity. Mutations in this gene have been associated with Holt-Oram syndrome, a developmental disorder affecting the heart and upper limbs. Several transcript variants encoding different isoforms have been described for this gene.
- **Genecards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=TBX5#summaries>

---

### 2. `IFITM1` — Interferon Induced Transmembrane Protein 1 

- **Fullname:** Interferon Induced Transmembrane Protein 1 
- **Functions:** nterferon-induced transmembrane (IFITM) proteins are a family of interferon induced antiviral proteins. The family contains five members, including IFITM1, IFITM2 and IFITM3 that belong to the CD225 superfamily. The protein encoded by this gene restricts cellular entry by diverse viral pathogens, such as influenza A virus, Ebola virus and Sars-CoV-2
- **GeneCards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=IFITM1#summaries>

---

### 3. `TNN` — Tenascin N

- **Full name:** Tenascin N 
- **Functions:** Predicted to enable integrin binding activity. Involved in positive regulation of sprouting angiogenesis; regulation of cell adhesion; and regulation of cell migration. Located in collagen-containing extracellular matrix. Part of tenascin complex.
Extracellular matrix protein that seems to be a ligand for ITGA8:ITGB1, ITGAV:ITGB1 and ITGA4:ITGB1 (By similarity) (. Involved in neurite outgrowth and cell migration in hippocampal explants (By similarity). During endochondral bone formation, inhibits proliferation and differentiation of proteoblasts mediated by canonical WNT signaling (By similarity). In tumors, stimulates angiogenesis by elongation, migration and sprouting of endothelial cells. Expressed in most mammary tumors, may facilitate tumorigenesis by supporting the migratory behavior of breast cancer cells 
- **GeneCards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=TNN#summaries>

---

### 4. `COL13A1` — Collagen Type XIII Alpha 1 Chain

- **Full name:** Collagen Type XIII Alpha 1 Chain
- **Functions:** This gene encodes the alpha chain of one of the nonfibrillar collagens. The function of this gene product is not known, however, it has been detected at low levels in all connective tissue-producing cells so it may serve a general function in connective tissues. Unlike most of the collagens, which are secreted into the extracellular matrix, collagen XIII contains a transmembrane domain and the protein has been localized to the plasma membrane. The transcripts for this gene undergo complex and extensive splicing involving at least eight exons. Like other collagens, collagen XIII is a trimer; it is not known whether this trimer is composed of one or more than one alpha chain isomer. A number of alternatively spliced transcript variants have been described, but the full length nature of some of them has not been determined
- **GeneCards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=COL13A1#summaries>

---

### 5. `IFITM3` — Interferon Induced Transmembrane Protein 3 

- **Full name:** Interferon Induced Transmembrane Protein 3 
- **Functions:** Interferon-induced transmembrane (IFITM) proteins are a family of interferon induced antiviral proteins. The family contains five members, including IFITM1, IFITM2 and IFITM3 and belong to the CD225 superfamily. The protein encoded by this gene restricts cellular entry by diverse viral pathogens, such as influenza A virus, Ebola virus and Sars-CoV-2. 
- **GeneCards:** <https://www.genecards.org/cgi-bin/carddisp.pl?gene=IFITM3#summaries>

---

# 10. Summary Statistics & Final Table

```{r summary-tables}
# ── Full summary of differentially expressed genes ───────────
cat("=== DIFFERENTIAL EXPRESSION SUMMARY ===\n")
cat(sprintf("Total genes analysed       : %d\n", nrow(rnaseq_clean)))
cat(sprintf("Upregulated   (FC>1, p<0.01): %d\n",
            sum(rnaseq_clean$Expression == "Upregulated")))
cat(sprintf("Downregulated (FC<-1, p<0.01): %d\n",
            sum(rnaseq_clean$Expression == "Downregulated")))
cat(sprintf("Not significant             : %d\n",
            sum(rnaseq_clean$Expression == "Not Significant")))

# ── Top 5 up and down in a combined table ────────────────────
cat("\n=== TOP 5 UPREGULATED ===\n")
print(head(upregulated %>%
             select(Gene, log2FoldChange, pvalue), 5))

cat("\n=== TOP 5 DOWNREGULATED ===\n")
print(head(downregulated %>%
             select(Gene, log2FoldChange, pvalue), 5))
```

---

# 11. Enhanced Volcano Plot with All Labels

```{r enhanced-volcano, fig.height=9}
# ── Final enhanced volcano with both up and down top genes ───
Volcano_plot_2 <-
  ggplot(rnaseq_clean,
       aes(x     = log2FoldChange,
           y     = neg_log10_pval,
           color = Expression)) +
  geom_point(alpha = 0.45, size = 1.4) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed", color = "grey20", linewidth = 0.6) +
  geom_hline(yintercept = -log10(0.01),
             linetype = "dashed", color = "grey20", linewidth = 0.6) +
  ggrepel::geom_label_repel(
    data        = top_genes,
    aes(label   = Gene),
    size        = 3.2,
    fontface    = "bold",
    fill        = "white",
    alpha       = 0.85,
    box.padding = 0.5,
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Upregulated"     = "#C62828",
      "Downregulated"   = "#1565C0",
      "Not Significant" = "grey70"
    )
  ) +
  # Annotate counts in each quadrant
  annotate("text",
           x =  max(rnaseq_clean$log2FoldChange, na.rm=TRUE) * 0.7,
           y =  max(rnaseq_clean$neg_log10_pval, na.rm=TRUE) * 0.92,
           label = paste0(sum(rnaseq_clean$Expression=="Upregulated"),
                          " Upregulated"),
           color = "#C62828", fontface = "bold", size = 4) +
  annotate("text",
           x =  min(rnaseq_clean$log2FoldChange, na.rm=TRUE) * 0.7,
           y =  max(rnaseq_clean$neg_log10_pval, na.rm=TRUE) * 0.92,
           label = paste0(sum(rnaseq_clean$Expression=="Downregulated"),
                          " Downregulated"),
           color = "#1565C0", fontface = "bold", size = 4) +
  labs(
    title    = "Volcano Plot — RNA-seq Differential Expression",
    subtitle = "Diseased Cell Line vs Diseased + Compound X  |  Top 5 Up & Down genes labelled",
    x        = "Log₂ Fold Change",
    y        = "-log₁₀(p-value)",
    color    = "Expression Status",
    caption  = "Thresholds: |Log2FC| > 1 and p-value < 0.01  |  Source: Turner (2015)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

Volcano_plot_2

ggsave("Volcano1.png", width = 12, height = 8, dpi = 150)
cat("✔ Saved: Volcano Plot.png\n")

```

---

> **Overall conclusion:**
> Compound X appears to act through a **dual mechanism** — simultaneously
> activating pro-apoptotic and structural repair pathways while suppressing
> tumour-promoting ECM remodelling and inflammatory signals. This pattern
> is consistent with a compound that **re-sensitises diseased cells to
> normal regulatory control** — a hallmark of effective therapeutic intervention.

---

*End of Task — Transcriptomics RNA-seq Differential Expression Analysis*
